---
title: "Babies named Nathan and Nate 1880-2014"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE, warning=FALSE, echo=FALSE}
##Loading packages for wrangling data

## install packages if necessary
list.of.packages <- c("dplyr", "ggplot2", 'plotly', 'usmap')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(dplyr)
library(ggplot2)
library(plotly)
library(usmap)
library(patchwork)
library(knitr)
library(ggrepel)

## set working directoy to wherever bab names csv is located
## this needs to be done in chunk preceding any chunks which rely on working directory setting 
knitr::opts_knit$set(root.dir = '~/Desktop/codes/babynames/')
```


### Getting names data

This data comes from [kaggle.com](https://www.kaggle.com/kaggle/us-baby-names), which has collected all recorded names for children born in the United States between the years 1880-2014.

```{r names-data, warning=FALSE, echo=TRUE}
## first read in baby names csv
## names <- read.csv('Baby-Name-Project/data/raw_data/NationalNames.csv')
## save as RDS and remove CSV to save space
names <- readRDS('Baby-Name-Project/data/raw_data/all-names.rds')

## reading in baby names by stat
## state <- read.csv('Baby-Name-Project/data/raw_data/StateNames.csv')
## save as RDS file and remove CSV
state <- readRDS('Baby-Name-Project/data/raw_data/state-names.rds')
```


## Top 10 boys names

Exploratory analysis of the top 10 boys names

```{r top 100 boys names, warning=FALSE, echo=TRUE}

## summing totals and organizing into descending order (most to least)
descending <- names %>%
  filter(Gender == 'M') %>%
  group_by(Name) %>%
  select(Name, Count) %>%
  summarise(total=sum(Count)) %>%
  arrange(desc(total))

## making table that displays top 10 names
top10 = descending[1:10,]

kable(top10, digits=0, 'html', caption = 'Top 10 Names for Baby Boys, US 1880-2014') %>% 
  kableExtra::kable_styling('striped', 'bordered') %>%
  kableExtra::footnote(general='Kaggle.com', general_title = 'Source: ', footnote_as_chunk = T)
```



## Top 10 girls names

Exploratory analysis of the top 10 girls names.

```{r top 100 girls names, warning=FALSE, echo=TRUE}

## summing totals and organizing into descending order (most to least)
descending <- names %>%
  filter(Gender == 'F') %>%
  group_by(Name) %>%
  select(Name, Count) %>%
  summarise(total=sum(Count)) %>%
  arrange(desc(total))

## making table that displays top 10 names
top10 = descending[1:10,]

kable(top10, digits=0, 'html', caption = 'Top 10 Names for Baby Girls, US 1880-2014') %>% 
  kableExtra::kable_styling('striped', 'bordered') %>%
  kableExtra::footnote(general='Kaggle.com', general_title = 'Source: ', footnote_as_chunk = T)
```



## Limiting to boys named Nathan and Nathan-realted Names

My main interest in this data set was to see how popular my own name is, specifically in comparison to popular variations.

```{r Nate-Nathan, warning=FALSE, echo=TRUE}

## filter out only babies with names 'Nathan' or 'Nate' and 'Male'
## this is accomplished using the filter() function available through dplyr package
dnn <- names %>% 
  filter(Gender == 'M', 
         Name == 'Nathan' | Name == 'Nate' | Name == 'Nathanial' | Name == 'Nathaniel' | Name == 'Nathanael')

## filtering out all other names for each state
state_dnn <- state %>% 
  filter(Gender == 'M', Name == 'Nathan')

## summing total number of Nathan's for each state
state_dnn_sum <- state_dnn %>% 
  group_by(State) %>%
  select(Name, state=State, Count) %>%
  summarize(total=sum(Count))

## getting total number of people per state
sum_names_state <- state %>%
  group_by(State) %>%
  select(state=State, Count) %>%
  summarize(total=sum(Count))

## getting proportion of Nathans for each state
## use set_colnames to alter header bc gets changed
prop_N <- data.frame(sum_names_state$state, (state_dnn_sum$total/sum_names_state$total)) %>%
  magrittr::set_colnames(c('state', 'total'))%>% 
  (function(x){
    df <- data_frame(state=x$state, prop = x$total*100)})

## summing total Nathaniels by state and filtering
state_dnnat <- state %>% 
  filter(Gender == 'M', Name == 'Nathaniel')

state_dnnat_sum <- state_dnnat %>% 
  group_by(State) %>%
  select(Name, state=State, Count) %>%
  summarize(total=sum(Count))

prop_Nat <- data.frame(sum_names_state$state, (state_dnnat_sum$total/sum_names_state$total)) %>%
  magrittr::set_colnames(c('state', 'total')) %>% 
  (function(x){
    df <- data_frame(state=x$state, prop = x$total*100)})

```



### Plot illustrating popularity of Nathan-related names

```{r plot, warning=FALSE, echo=TRUE}

## cleaner looking graph with annotations
p1 <- ggplot(data=dnn, aes(x=Year, y=Count, color=Name)) + 
  geom_line(size=1) + 
  labs(title='Baby boys with Nathan-related names',
       subtitle='United States, 1880-2014', 
       caption='Source: www.kaggle.com') + 
  ylab('Number of babies') +
  scale_x_continuous(breaks=seq(1880,2014, by=10)) +
  annotate('rect', xmin=1989, xmax=1991, ymin=0, ymax=Inf, fill= 'cadetblue3', alpha=0.6) + 
  annotate('text', label = '1990', y=12500, x=1988, size=5, hjust='right') + 
  theme(
    plot.title = element_text(face='bold', size = 16),
    plot.subtitle = element_text(size=13),
    plot.margin = unit(c(1,1,1,1), 'lines'),
    axis.text = element_text(size=10, color='black'),
    axis.title.y = element_text(size=12, face='bold', margin = margin(t=0,r=10,b=0,l=0)),
    axis.title.x = element_text(size=12, face='bold', margin = margin(t=10,r=0,b=0,l=0)),
    legend.position = c(0.15,0.7),
    legend.text = element_text(size=12),
    legend.title = element_text(size=12, face='bold'),
    legend.background = element_rect(fill='white', size=0.5, linetype='solid', color='black')
  )

p1

```

There are a few interesting takeaways from this graph. First, Nathan is by far the most popular variation. However, there is a several decade span in which Nathaniel is more popular. Throughout the 1910s, Nathan was the variation of choice. However, beginning ~1920 and lasting throughout most of the 1950s, Nathaniel notched the #1 position. Then ~1958, Nathan reclaimed the most popular variation, a trend which has lasted at least through 2014.

In terms of rises and falls in popularity, Nathan and Nathaniel exhibit a similar trend (not all that suprising). You would expect to see this kind of relationship among name variants. Interestingly, other observed variations (Nathanial, Nathanael, Nate) do not mirror those trends.



### Interactive Plot

Basically the same graph as above, except that it has an interactive component.

```{r interactive-plot, warning=FALSE, echo=TRUE}

## interactive plot
p2 <- ggplot(data=dnn, aes(x=Year, y=Count, color=Name)) + 
  geom_line() + 
  geom_point() + 
  ylab('Number of babies') + 
  theme(
    plot.margin = unit(c(1,1,1,1), 'lines')
  )

## in order to make interactive, we will view the plot using ggplotly() function.
ggplotly(p2)

## you should be able to hover mouse over individual points to see count and year
## if you don't want the graph to show up inside R markdown file:
    ## click on gear aside of knit at top
    ## select 'Chuck Output in Console'
```



### Mapping total number of Nathans by state

Making a heatplot for proportion of Nathans born in each state (shown as percentage relative to the counts of all other names). For example, say a state has a value of 0.4%. This means that 0.4% of all people born in that state since 1880 were named Nathan. That's a frequency of 4 in 1000 people.

```{r Nathans by state, warning=FALSE, echo=TRUE}

p3 <- plot_usmap(data=prop_N, values = 'prop') + 
  scale_fill_gradient(name = 'Percentage', low='blue', high='red') + 
  labs(title='Percentage of Babies Named Nathan By State', 
       subtitle = 'US babies born 1880-2014', 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(size=11, face='bold'),
    legend.text = element_text(size=9),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )

p3
```

The major take aways from this graph: Nathan is most popular in Utah (intestingly) and least popular in New Jersey.



## Mapping total Nathaniels by State

Same principle as the above graph but using Nathaniel instead of Nathan.

```{r Nathaniels by state, warning=FALSE, echo=TRUE}

p4 <- plot_usmap(data=prop_Nat, values = 'prop') + 
  scale_fill_gradient(name = 'Percentage', low='blue', high='red') + 
  labs(title='Percentage of Babies Named Nathaniel By State', 
       subtitle = 'US babies born 1880-2014', 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(size=11, face='bold'),
    legend.text = element_text(size=9),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )

p4

```

The major takeaways from this graph: Nathaniel is most popular in South Carolina and least popular in North Dakota. However, if we take a look at the scale bar for this graph and compare it to the one for the Nathan-By-State graph, it's clear that Nathan is the more popular name. 

**Note**: Remember, this is for the entire dataset. There are likely to be local trends (e.g., decade-specific trends). See the Nathan-related names graph above for reference.



## State preference for Nathan or Nathaniel

```{r state preference, warning=FALSE, echo=TRUE}

## combining Nathan and Nathaniel proportions into single data frame
props <- data.frame(prop_N$state, prop_N$prop, prop_Nat$prop)
colnames(props) <- c('state', 'Nathan', 'Nathaniel')

## writing function that will determine which is more popular for each state
## uses numeric representation: 1==Nathan, 2==Nathaniel
nathan_or_nathaniel <- function(nathan,nathaniel){
  bigger = c()
  for (val in 1:length(nathan)){
    if (nathan[val] > nathaniel[val]){
      bigger = append(bigger, '1')
    }else{
      bigger = append(bigger, '2')
    }
  }
  return(bigger)
}

## running function and adding as new column to props data frame
non <- nathan_or_nathaniel(props$Nathan, props$Nathaniel)
props$non <- non

## determining which is more popular by state
det_nvn <- function(non){
  one = c()
  two = c()
  for (val in non){
    if (val == '1'){
      one = append(one, 1)
    }else{
      two = append(two, 1)
    }
  }
  if (length(one) > length(two)){
    return(paste0('Majority of states (including DC) prefer Nathan: ', length(one), ' states'))
  }else{
    return(paste0('Majority of states (including DC) prefer Nathaniel: ', length(two), ' states'))
  }
}

which_is_more <- det_nvn(props$non)

## plotting by state
p_non <- plot_usmap(data=props, values = 'non') + 
  scale_fill_manual(name='Variation', 
                    labels = c('Nathan', 'Nathaniel'), 
                    values=c('Peach Puff', 'Tomato')) + 
  labs(title='Nathan vs Nathaniel By State', 
       subtitle = paste0('US babies born 1880-2014\n\n', which_is_more), 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(face='bold', size=13),
    legend.text = element_text(size=11),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )
p_non

```

This figure provides a nice visulation of what we already could've guessed from the two heatmaps above. Nathan is the more popular variation for an overwhelming majority of states. The only two exceptions are South Carolina (0.17% v 0.12%) and Washington DC (0.14% v 0.11%).



```{r saving data, warning=FALSE, echo=FALSE}
## raw data as .rds
## it is good pratice to save imported data in event it gets lost or corrupt. 
## always save anthing obtained from online (especially from html)
saveRDS(names, 'Baby-Name-Project/data/raw_data/all-names.rds')

## processed data containing only Nathan and Nathan-realted names for boys
saveRDS(dnn, 'Baby-Name-Project/data/processed_data/Nathan-related-names.rds')

## state data unprocessed
saveRDS(state, 'Baby-Name-Project/data/raw_data/state-names.rds')

## processed data containing only Nathan for each state, including years
saveRDS(state_dnn, 'Baby-Name-Project/data/processed_data/state-Nathans-by-year.rds')

## processed data containing summed Nathans by state
## meaning, only one value for each state which is all the Nathans that were ever born there
saveRDS(state_dnn_sum, 'Baby-Name-Project/data/processed_data/state-Nathans-total.rds')

## processed data containing total number of babies per state
saveRDS(sum_names_state, 'Baby-Name-Project/data/processed_data/total-babies-state.rds')

## normalized data for percentages of Nathans by state
saveRDS(prop_N, 'Baby-Name-Project/data/processed_data/prop-Nathans-state.rds')

## processed data containing only Nathan for each state, including years
saveRDS(state_dnnat, 'Baby-Name-Project/data/processed_data/state-Nathaniels-by-year.rds')

## processed data containing summed Nathniels by state
saveRDS(state_dnnat_sum, 'Baby-Name-Project/data/processed_data/state-Nathaniels-total.rds')

## normalized data for percentages of Nathans by state
saveRDS(prop_Nat, 'Baby-Name-Project/data/processed_data/prop-Nathaniels-state.rds')

```


```{r saving image, warning=FALSE, echo=FALSE}

ggsave(plot=p1, filename='Nathan-Names.png', 
       path = 'Baby-Name-Project/figures/explanatory_figures/', 
       height=5.58, width = 8, unit=c('in'))

ggsave(plot=p3, filename='Nathan-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

ggsave(plot=p4, filename='Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

## combined Nathan and Nathaniel by state
p5=p3+p4+plot_layout(ncol=1)

ggsave(plot=p5, filename='Nathan-and-Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=8, width=6, unit=c('in'))

ggsave(plot=p_non, filename='Nathan-vs-Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

```