---
title: "Babies named Nathan and Nate 1880-2014"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE, warning=FALSE, echo=FALSE}
##Loading packages for wrangling data

## install packages if necessary
list.of.packages <- c("dplyr", "ggplot2", 'plotly', 'usmap')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(dplyr)
library(ggplot2)
library(plotly)
library(usmap)
library(patchwork)
library(knitr)
library(ggrepel)

## set working directoy to wherever bab names csv is located
## this needs to be done in chunk preceding any chunks which rely on working directory setting 
knitr::opts_knit$set(root.dir = '~/Desktop/codes/babynames/')
```

### Getting names data

This data comes from [kaggle.com](https://www.kaggle.com/kaggle/us-baby-names), which has collected all recorded names for children born in the United States between the years 1880-2014.

```{r names-data, warning=FALSE, echo=TRUE}
## first read in baby names csv
## names <- read.csv('Baby-Name-Project/data/raw_data/NationalNames.csv')
## save as RDS and remove CSV to save space
names <- readRDS('Baby-Name-Project/data/raw_data/all-names.rds')

## reading in baby names by stat
## state <- read.csv('Baby-Name-Project/data/raw_data/StateNames.csv')
## save as RDS file and remove CSV
state <- readRDS('Baby-Name-Project/data/raw_data/state-names.rds')
```


## Top 10 boys names

```{r top 100 boys names, warning=FALSE, echo=TRUE}

## summing totals and organizing into descending order (most to least)
descending <- names %>%
  filter(Gender == 'M') %>%
  group_by(Name) %>%
  select(Name, Count) %>%
  summarise(total=sum(Count)) %>%
  arrange(desc(total))

## making table that displays top 10 names
top10 = descending[1:10,]

kable(top10, digits=0, 'html', caption = 'Top 10 Names for Baby Boys, US 1880-2014') %>% 
  kableExtra::kable_styling('striped', 'bordered') %>%
  kableExtra::footnote(general='Kaggle.com', general_title = 'Source: ', footnote_as_chunk = T)
```

## Limiting to boys named Nathan and Nathan-realted Names

```{r Nate-Nathan, warning=FALSE, echo=TRUE}

## filter out only babies with names 'Nathan' or 'Nate' and 'Male'
## this is accomplished using the filter() function available through dplyr package
dnn <- names %>% 
  filter(Gender == 'M', 
         Name == 'Nathan' | Name == 'Nate' | Name == 'Nathanial' | Name == 'Nathaniel' | Name == 'Nathanael')

## filtering out all other names for each state
state_dnn <- state %>% 
  filter(Gender == 'M', Name == 'Nathan')

## summing total number of Nathan's for each state
state_dnn_sum <- state_dnn %>% 
  group_by(State) %>%
  select(Name, state=State, Count) %>%
  summarize(total=sum(Count))

## getting total number of people per state
sum_names_state <- state %>%
  group_by(State) %>%
  select(state=State, Count) %>%
  summarize(total=sum(Count))

## getting proportion of Nathans for each state
## use set_colnames to alter header bc gets changed
prop_N <- data.frame(sum_names_state$state, (state_dnn_sum$total/sum_names_state$total)) %>%
  magrittr::set_colnames(c('state', 'total'))%>% 
  (function(x){
    df <- data_frame(state=x$state, prop = x$total*100)})

## summing total Nathaniels by state and filtering
state_dnnat <- state %>% 
  filter(Gender == 'M', Name == 'Nathaniel')

state_dnnat_sum <- state_dnnat %>% 
  group_by(State) %>%
  select(Name, state=State, Count) %>%
  summarize(total=sum(Count))

prop_Nat <- data.frame(sum_names_state$state, (state_dnnat_sum$total/sum_names_state$total)) %>%
  magrittr::set_colnames(c('state', 'total')) %>% 
  (function(x){
    df <- data_frame(state=x$state, prop = x$total*100)})

```


### Creating interactive plot

```{r plot, warning=FALSE, echo=TRUE}

## cleaner looking graph with annotations
p1 <- ggplot(data=dnn, aes(x=Year, y=Count, color=Name)) + 
  geom_line(size=1) + 
  labs(title='Baby boys with Nathan-related names',
       subtitle='United States, 1880-2014', 
       caption='Source: www.kaggle.com') + 
  ylab('Number of babies') +
  scale_x_continuous(breaks=seq(1880,2014, by=10)) +
  annotate('rect', xmin=1989, xmax=1991, ymin=0, ymax=Inf, fill= 'cadetblue3', alpha=0.6) + 
  annotate('text', label = '1990', y=12500, x=1988, size=5, hjust='right') + 
  theme(
    plot.title = element_text(face='bold', size = 16),
    plot.subtitle = element_text(size=13),
    plot.margin = unit(c(1,1,1,1), 'lines'),
    axis.text = element_text(size=10, color='black'),
    axis.title.y = element_text(size=12, face='bold', margin = margin(t=0,r=10,b=0,l=0)),
    axis.title.x = element_text(size=12, face='bold', margin = margin(t=10,r=0,b=0,l=0)),
    legend.position = c(0.15,0.7),
    legend.text = element_text(size=12),
    legend.title = element_text(size=12, face='bold'),
    legend.background = element_rect(fill='white', size=0.5, linetype='solid', color='black')
  )

p1

```

### Interactive Plot

```{r interactive-plot, warning=FALSE, echo=TRUE}

## interactive plot
p2 <- ggplot(data=dnn, aes(x=Year, y=Count, color=Name)) + 
  geom_line() + 
  geom_point() + 
  ylab('Number of babies') + 
  theme(
    plot.margin = unit(c(1,1,1,1), 'lines')
  )

## in order to make interactive, we will view the plot using ggplotly() function.
ggplotly(p2)

## you should be able to hover mouse over individual points to see count and year
## if you don't want the graph to show up inside R markdown file:
    ## click on gear aside of knit at top
    ## select 'Chuck Output in Console'
```


### Mapping total number of Nathans by state

Making a heatplot for proportion of Nathans born in each state

```{r Nathans by state, warning=FALSE, echo=TRUE}

p3 <- plot_usmap(data=prop_N, values = 'prop') + 
  scale_fill_gradient(name = 'Percentage', low='blue', high='red') + 
  labs(title='Percentage of Babies Named Nathan By State', 
       subtitle = 'US babies born 1880-2014', 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(size=11, face='bold'),
    legend.text = element_text(size=9),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )

p3
```

## Mapping total Nathaniels by State

```{r Nathaniels by state, warning=FALSE, echo=TRUE}

p4 <- plot_usmap(data=prop_Nat, values = 'prop') + 
  scale_fill_gradient(name = 'Percentage', low='blue', high='red') + 
  labs(title='Percentage of Babies Named Nathaniel By State', 
       subtitle = 'US babies born 1880-2014', 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(size=11, face='bold'),
    legend.text = element_text(size=9),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )

p4

```

## State preference for Nathan or Nathaniel

```{r state preference, warning=FALSE, echo=TRUE}

## combining Nathan and Nathaniel proportions into single data frame
props <- data.frame(prop_N$state, prop_N$prop, prop_Nat$prop)
colnames(props) <- c('state', 'Nathan', 'Nathaniel')

## writing function that will determine which is more popular for each state
## uses numeric representation: 1==Nathan, 2==Nathaniel
nathan_or_nathaniel <- function(nathan,nathaniel){
  bigger = c()
  for (val in 1:length(nathan)){
    if (nathan[val] > nathaniel[val]){
      bigger = append(bigger, '1')
    }else{
      bigger = append(bigger, '2')
    }
  }
  return(bigger)
}

## running function and adding as new column to props data frame
non <- nathan_or_nathaniel(props$Nathan, props$Nathaniel)
props$non <- non

## determining which is more popular by state
det_nvn <- function(non){
  one = c()
  two = c()
  for (val in non){
    if (val == '1'){
      one = append(one, 1)
    }else{
      two = append(two, 1)
    }
  }
  if (length(one) > length(two)){
    diff = (length(one) + 1) - length(two)
    return(paste0('Nathan by ', diff, ' states (including DC)'))
  }else{
    diff = (length(two) + 1) - length(one)
    return(paste0('Nathaniel by ', diff, ' states (including DC'))
  }
}

which_is_more <- det_nvn(props$non)

## plotting by state
p_non <- plot_usmap(data=props, values = 'non') + 
  scale_fill_manual(name='Variation', 
                    labels = c('Nathan', 'Nathaniel'), 
                    values=c('Peach Puff', 'Tomato')) + 
  labs(title='Nathan vs Nathaniel By State', 
       subtitle = paste0('US babies born 1880-2014\n\n', which_is_more), 
       caption='Source: Kaggle.com') + 
  theme(
    legend.position = 'right',
    legend.title = element_text(face='bold', size=13),
    legend.text = element_text(size=11),
    plot.title = element_text(size=16, face='bold'),
    plot.subtitle = element_text(size=13),
    plot.caption = element_text(size=9)
  )
p_non

```

```{r saving data, warning=FALSE, echo=FALSE}
## raw data as .rds
## it is good pratice to save imported data in event it gets lost or corrupt. 
## always save anthing obtained from online (especially from html)
saveRDS(names, 'Baby-Name-Project/data/raw_data/all-names.rds')

## processed data containing only Nathan and Nathan-realted names for boys
saveRDS(dnn, 'Baby-Name-Project/data/processed_data/Nathan-related-names.rds')

## state data unprocessed
saveRDS(state, 'Baby-Name-Project/data/raw_data/state-names.rds')

## processed data containing only Nathan for each state, including years
saveRDS(state_dnn, 'Baby-Name-Project/data/processed_data/state-Nathans-by-year.rds')

## processed data containing summed Nathans by state
## meaning, only one value for each state which is all the Nathans that were ever born there
saveRDS(state_dnn_sum, 'Baby-Name-Project/data/processed_data/state-Nathans-total.rds')

## processed data containing total number of babies per state
saveRDS(sum_names_state, 'Baby-Name-Project/data/processed_data/total-babies-state.rds')

## normalized data for percentages of Nathans by state
saveRDS(prop_N, 'Baby-Name-Project/data/processed_data/prop-Nathans-state.rds')

## processed data containing only Nathan for each state, including years
saveRDS(state_dnnat, 'Baby-Name-Project/data/processed_data/state-Nathaniels-by-year.rds')

## processed data containing summed Nathniels by state
saveRDS(state_dnnat_sum, 'Baby-Name-Project/data/processed_data/state-Nathaniels-total.rds')

## normalized data for percentages of Nathans by state
saveRDS(prop_Nat, 'Baby-Name-Project/data/processed_data/prop-Nathaniels-state.rds')

```


```{r saving image, warning=FALSE, echo=FALSE}

ggsave(plot=p1, filename='Nathan-Names.png', 
       path = 'Baby-Name-Project/figures/explanatory_figures/', 
       height=5.58, width = 8, unit=c('in'))

ggsave(plot=p3, filename='Nathan-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

ggsave(plot=p4, filename='Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

## combined Nathan and Nathaniel by state
p5=p3+p4+plot_layout(ncol=1)

ggsave(plot=p5, filename='Nathan-and-Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=8, width=6, unit=c('in'))

ggsave(plot=p_non, filename='Nathan-vs-Nathaniel-By-State.png',
       path='Baby-Name-Project/figures/explanatory_figures/',
       height=4, width=6, unit=c('in'))

```